# Basic Hypothesis testing

### **1. Data Loading and Preprocessing**

```{r}
# Load required libraries
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
# library(forecast)
library(mgcv)
library(corrplot)
library(ggpubr)
library(cluster)
library(factoextra)

# Load data
yearly <- read_excel("./01-analysis/data/Area_RF_Sea.xlsx", sheet = "Area_Yearly")
monthly <- read_excel("./01-analysis/data/Area_RF_Sea.xlsx", sheet = "Area_Monthly")

# Preprocess monthly data: Create a Date column
monthly <- monthly %>%
  mutate(Date = as.Date(paste(year, month, "01", sep = "-"), format = "%Y-%m-%d"))
```
---

### **2. Yearly Data Analysis**
#### **Hypothesis 1: Water area has changed significantly over time**
```{r}
yearly_long <- yearly %>%
  pivot_longer(
    cols = starts_with("wa_"),
    names_to = "Classifier",
    values_to = "Area"
  ) %>%
  mutate(
    Classifier = gsub("wa_", "", Classifier),
    Classifier = factor(Classifier)
  )

# Professional color palette (colorblind-friendly + print-friendly)
line_colors <- c("#1a476f", "#90353b", "#55752f", "#e37e00", "#6e8e84")

ggplot(yearly_long, aes(x = year, y = Area,
                       color = Classifier)) +
  geom_line(linewidth = 0.8) +
  geom_point(
    size = 3,
    aes(),
    show.legend = FALSE
  ) +
  labs(
    x = "Year",
    y = expression(Water~Area~(ha))
  ) +
  scale_x_continuous(
    breaks = seq(min(yearly_long$year), max(yearly_long$year), by = 2),
    limits = c(min(yearly_long$year), max(yearly_long$year)),
    expand = expansion(mult = 0.02)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.05, 0.1))
  ) +
  scale_color_manual(values = line_colors) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8)) +  # Different point shapes
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key.width = unit(1.5, "cm"),  # Wider legend keys
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.2),
    axis.line = element_line(color = "black"),
    panel.border = element_rect(fill = NA, color = "black", linewidth = 0.5),
    plot.margin = unit(c(5,5,5,5), "mm"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "black")
  )

# Save with optimal settings
ggsave("yearly_wa_trend.pdf",
       width = 10,
       height = 8,
       units = "in",
       device = cairo_pdf)  # Better font handling
```

**Figure 1**: All classifiers show variability, but RF (Random Forest) exhibits a slight decline.
**Interpretation**: Visual trends suggest changes, but statistical validation is needed.

**Statistical Test**:
```{r}
# Mann-Kendall trend tests for all water area classifiers
if (!require(trend)) { install.packages('trend'); library(trend)}

# CART water area ---------------------------------------------
mk_test_cart_yr <- mk.test(yearly$wa_cart_yr_ha)
cat("\nMann-Kendall Trend Test for CART Water Area Yearly:\n")
print(mk_test_cart_yr)

mk_test_cart <- mk.test(monthly$wa_cart_ha)
cat("\nMann-Kendall Trend Test for CART Water Area Monthly:\n")
print(mk_test_cart)


# MD water area---------------------------------------------
mk_test_md_yr <- mk.test(yearly$wa_md_yr_ha)
cat("\nMann-Kendall Trend Test for MD Water Area Yearly:\n")
print(mk_test_md_yr)

mk_test_md <- mk.test(monthly$wa_md_ha)
cat("\nMann-Kendall Trend Test for MD Water Area Monthly:\n")
print(mk_test_md)


# NB water area---------------------------------------------
mk_test_nb_yr <- mk.test(yearly$wa_nb_yr_ha)
cat("\nMann-Kendall Trend Test for NB Water Area Yearly:\n")
print(mk_test_nb_yr)

mk_test_nb <- mk.test(monthly$wa_nb_ha)
cat("\nMann-Kendall Trend Test for NB Water Area Monthly:\n")
print(mk_test_nb)

# RF water area ---------------------------------------------
mk_test_rf_yr <- mk.test(yearly$wa_rf_yr_ha)
cat("\nMann-Kendall Trend Test for RF Water Area Yearly:\n")
print(mk_test_rf_yr)

mk_test_rf <- mk.test(monthly$wa_rf_ha)
cat("\nMann-Kendall Trend Test for RF Water Area Monthly:\n")
print(mk_test_rf)

```
**Output**

### **Interpreting the Mann-Kendall Trend Test Output**

The **Mann-Kendall (MK) test** is a non-parametric method used to detect trends in time-series data. Below is an explanation of the key components of the output for each classifier (CART, MD, NB, and RF):

#### **C) `tau` (Kendall's Tau Coefficient)**
- Measures **trend strength and direction**:
  - **Range**: `-1` (perfect decreasing trend) to `+1` (perfect increasing trend)
  - **`tau ≈ 0`** → No trend
  - **`tau > 0`** → Increasing trend
  - **`tau < 0`** → Decreasing trend

#### **D) `S` (Kendall's S Statistic)**
- Sum of score differences between all pairs of data points.
  - **Higher `|S|`** → Stronger trend
  - **`S = 0`** → No trend

### **3. Interpretation of Your Results**
#### **a) CART Water Area (`p = 0.2714`)**
- **`z = 1.0997`** (slight positive trend)
- **`p = 0.2714`** (**Not significant**, `p > 0.05`)
- **`tau = 0.2595`** (weak increasing trend)
- **Conclusion**: **No significant trend**

#### **b) MD Water Area (`p = 0.04674`)**
- **`z = 1.9886`** (moderate positive trend)
- **`p = 0.04674`** (**Significant**, `p < 0.05`)
- **`tau = 0.4545`** (moderate increasing trend)
- **Conclusion**: **Significant increasing trend**

#### **c) NB Water Area (`p = 0.003192`)**
- **`z = 2.9486`** (strong positive trend)
- **`p = 0.003192`** (**Highly significant**, `p < 0.01`)
- **`tau = 0.6667`** (strong increasing trend)
- **Conclusion**: **Highly significant increasing trend**

#### **d) RF Water Area (`p = 0.3704`)**
- **`z = 0.89566`** (slight positive trend)
- **`p = 0.3704`** (**Not significant**, `p > 0.05`)
- **`tau = 0.2154`** (weak increasing trend)
- **Conclusion**: **No significant trend**

---

### **4. Summary Table**
| Classifier | `z`-value | `p`-value | Kendall's Tau (`τ`) | Trend Significance |
|------------|----------|----------|------------------|------------------|
| **CART**   | 1.0997   | 0.2714   | 0.2595           | Not significant  |
| **MD**     | 1.9886   | **0.04674** | 0.4545         | **Significant ↑** |
| **NB**     | 2.9486   | **0.003192** | 0.6667       | **Highly Significant ↑** |
| **RF**     | 0.89566  | 0.3704   | 0.2154           | Not significant  |

### **5. Key Takeaways**
- **MD and NB** show **significant increasing trends** in water area.
- **CART and RF** show **no significant trends**.
- **NB has the strongest trend** (`τ = 0.6667`, `p < 0.01`).

---

### **3. Monthly Data Analysis**

**Statistical Test**:
```{r}
# ANOVA for seasonal (monthly) differences in water area
anova_cart <- aov(wa_cart_ha ~ Month, data = monthly)
anova_md   <- aov(wa_md_ha ~ Month, data = monthly)
anova_nb   <- aov(wa_nb_ha ~ Month, data = monthly)
anova_rf   <- aov(wa_rf_ha ~ Month, data = monthly)

# Summarize results
cat("\nANOVA for CART Water Area by Month:\n")
summary(anova_cart)

cat("\nANOVA for MD Water Area by Month:\n")
summary(anova_md)

cat("\nANOVA for NB Water Area by Month:\n")
summary(anova_nb)

cat("\nANOVA for RF Water Area by Month:\n")
summary(anova_rf)

TukeyHSD(anova_cart)
boxplot(wa_cart_ha ~ Month, data = monthly, main = "CART Water Area by Month")
```

**Interpretation of the ANOVA Output for Seasonal (Monthly) Differences**

ANOVA (Analysis of Variance) tests whether there are **statistically significant differences** in water area across different months for each classifier (CART, MD, NB, RF). Here’s how to interpret the results:


**D) F-statistic (`F value`)**
- **F = `Month Mean Sq / Residuals Mean Sq`**
  - Tests if **monthly variation is larger than random noise**.
  - **Higher F = Stronger evidence of seasonal differences**.

**E) p-value (`Pr(>F)`)**
- **Determines statistical significance**:
  - **p < 0.05 → Significant seasonal differences.**
  - **p ≥ 0.05 → No significant differences.**
- **Significance codes**:
  - `***` (p < 0.001), `**` (p < 0.01), `*` (p < 0.05), `.` (p < 0.1).


### **4. Summary Table**
| Classifier | F-value | p-value       | Significance | Conclusion                     |
|------------|---------|---------------|--------------|--------------------------------|
| **CART**   | 6.371   | **2.13e-08**  | ***          | Extremely strong seasonality   |
| **MD**     | 5.679   | **1.94e-07**  | ***          | Very strong seasonality        |
| **NB**     | 4.528   | **8.36e-06**  | ***          | Strong seasonality             |
| **RF**     | 2.821   | **0.00244**   | **           | Moderate seasonality           |



#### **Hypothesis 4: Classifiers agree on water area estimates: Bland-Altman Plot**
1. Purpose of the Plots
Bland-Altman plots assess agreement between two measurement methods (in this case, classifiers). They answer:

Do the classifiers systematically over/underestimate water area compared to each other?

How large are the discrepancies between classifiers?

Are disagreements consistent across all water area magnitudes?

2. Key Components of Each Subplot
X-axis: Mean of two classifiers
Represents the "average truth" between the two methods.
Y-axis: Difference between two classifiers
Positive values → First classifier overestimates relative to the second.
Negative values → First classifier underestimates.
Red horizontal line: Mean bias (average difference).
Dashed line: Perfect agreement (y = 0).

```{r}
library(ggplot2)
library(patchwork)

# Calculate global y-axis limits for all differences
cart_md_diff <- monthly$wa_cart_ha - monthly$wa_md_ha
cart_nb_diff <- monthly$wa_cart_ha - monthly$wa_nb_ha
cart_rf_diff <- monthly$wa_cart_ha - monthly$wa_rf_ha
md_nb_diff <- monthly$wa_md_ha - monthly$wa_nb_ha
md_rf_diff <- monthly$wa_md_ha - monthly$wa_rf_ha
nb_rf_diff <- monthly$wa_nb_ha - monthly$wa_rf_ha

all_diffs <- c(cart_md_diff, cart_nb_diff, cart_rf_diff,
               md_nb_diff, md_rf_diff, nb_rf_diff)

# Get symmetric limits based on maximum absolute difference
max_abs <- ceiling(max(abs(all_diffs)))
y_limits <- c(-max_abs, max_abs)

# Create plots with identical y-axis scales
create_ba_plot <- function(x1, x2, title) {
  differences <- x1 - x2
  mean_diff <- mean(differences)

  ggplot(monthly, aes(x = (x1 + x2)/2, y = differences)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = mean_diff, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = title, x = "Mean (ha)", y = "Difference (ha)") +
    ylim(y_limits) +  # Enforce global y-axis limits
    theme_minimal(base_size = 10)
}

# Generate all plots
ba1 <- create_ba_plot(monthly$wa_cart_ha, monthly$wa_md_ha, "CART vs MD")
ba2 <- create_ba_plot(monthly$wa_cart_ha, monthly$wa_nb_ha, "CART vs NB")
ba3 <- create_ba_plot(monthly$wa_cart_ha, monthly$wa_rf_ha, "CART vs RF")
ba4 <- create_ba_plot(monthly$wa_md_ha, monthly$wa_nb_ha, "MD vs NB")
ba5 <- create_ba_plot(monthly$wa_md_ha, monthly$wa_rf_ha, "MD vs RF")
ba6 <- create_ba_plot(monthly$wa_nb_ha, monthly$wa_rf_ha, "NB vs RF")

# Combine plots
combined_plot <- (ba1 + ba2 + ba3) / (ba4 + ba5 + ba6) +
  plot_annotation(
    title = "Bland-Altman Plots: Pairwise Comparison of Water Area Classifiers",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))

# Display the combined plot
ggsave("Bland-Altman_combined_plot.pdf")
```
