# Load required packages

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(readxl, lubridate, ggplot2, dplyr, tidyr, corrplot, forecast, ggcorrplot, Hmisc, tsibble, feasts, patchwork, readxl, zoo, fable, tseries)
```

# Import data
```{r}
library(lubridate)
monthly_data <- read_excel("./01-analysis/data/Area_RF_Sea.xlsx", sheet = "Area_Monthly")

# Preprocess
df <- monthly_data %>%
  mutate(
    date = yearmonth(make_date(year, month)),  # Convert to year-month type
    season = factor(seasons, levels = c("Winter", "Summer", "Monsoon"))
  ) %>%
  select(-index, -seasons) %>%
  relocate(date) %>%
  filter(wa_rf_ha > 0) %>%  # Remove zero-water-area observations
  distinct(date, .keep_all = TRUE) %>%  # Ensure unique dates
  arrange(date)

# Create monthly time series
ts_data <- df %>%
  as_tsibble(index = date) %>%  # No need for `interval` argument
  fill_gaps() %>%  # Fill missing months
  mutate(across(where(is.numeric), zoo::na.locf, na.rm = FALSE))

print(ts_data)
```

## 1. Correlation analysis
```{r}
cor_vars <- df %>% select(wa_cart_ha, wa_md_ha, wa_nb_ha, wa_rf_ha, precipitation_mm_mon, Temperature)

# Correlation matrix with significance
cor_matrix <- Hmisc::rcorr(as.matrix(cor_vars))

# Visualization: Correlation plot
library(ggcorrplot)
cor_plot <- ggcorrplot(cor_matrix$r, hc.order = TRUE, type = "lower", lab = TRUE, p.mat = cor_matrix$P, insig = "blank") + theme_minimal(base_size = 12) + labs(title = "Correlation Matrix of Monthly Variables")

ggsave("Correlation_monthly_all.pdf")
```

# 2. Time series decomposition
```{r}
library(fable)
library(tsibble)
library(feasts)  # For autoplot

decomp_plot_cart <- ts_data %>%
  model(STL(wa_cart_ha ~ season(window = 13), robust = TRUE)) %>%
  components() %>%                   # Extract decomposition components
  autoplot() +                       # Plot components
  theme_bw(base_size = 12) +
  labs(title = "STL Decomposition of CART Water Area")

decomp_plot_md <- ts_data %>%
  model(STL(wa_md_ha ~ season(window = 13))) %>%  # Use fable::STL
  components() %>%                   # Extract decomposition components
  autoplot() +                       # Plot components
  theme_bw(base_size = 12) +
  labs(title = "STL Decomposition of MD Water Area")

decomp_plot_nb <- ts_data %>%
  model(STL(wa_nb_ha ~ season(window = 13))) %>%  # Use fable::STL
  components() %>%                   # Extract decomposition components
  autoplot() +                       # Plot components
  theme_bw(base_size = 12) +
  labs(title = "STL Decomposition of NB Water Area")

decomp_plot_rf <- ts_data %>%
  model(STL(wa_rf_ha ~ season(window = 13))) %>%  # Use fable::STL
  components() %>%                   # Extract decomposition components
  autoplot() +                       # Plot components
  theme_bw(base_size = 12) +
  labs(title = "STL Decomposition of RF Water Area")

combined_plot <- ((decomp_plot_cart | decomp_plot_md) / (decomp_plot_nb | decomp_plot_rf) ) + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(face = "bold"))

# ggsave("combined_analysis.pdf", combined_plot,  width = 16, height = 12, dpi = 300)

ggsave("combined_analysis.png", combined_plot,  width = 16, height = 12, dpi = 300)
```
## Process the data
```{r}
data <- monthly_data
data$date <- as.Date(paste(data$year, data$month, "01", sep = "-"))

ts_data <- data %>%
  mutate(
    wa_cart_ha = ts(wa_cart_ha, start = c(2013, 1), frequency = 12),
    wa_md_ha = ts(wa_md_ha, start = c(2013, 1), frequency = 12),
    wa_nb_ha = ts(wa_nb_ha, start = c(2013, 1), frequency = 12),
    wa_rf_ha = ts(wa_rf_ha, start = c(2013, 1), frequency = 12),
    precipitation_mm_mon = ts(precipitation_mm_mon, start = c(2013, 1), frequency = 12),
    Temperature = ts(Temperature, start = c(2013, 1), frequency = 12)
  )
```

## 3.  Autocorrelation Analysis
Perform autocorrelation and partial autocorrelation for all four classifiers.

```{r}
generate_acf_pacf <- function(variable, title_prefix) {
  # ACF Plot (unchanged from your version)
  acf_data <- Acf(ts_data[[variable]], lag.max = 36, plot = FALSE)
  acf_df <- with(acf_data, data.frame(lag = lag, ACF = acf))

  acf_plot <- ggplot(acf_df, aes(x = lag, y = ACF)) +
    geom_bar(stat = "identity", position = "identity",
             width = 0.2, fill = "#3f3f3f", color = NA) +
    geom_point(size = 3, color = "#8a8a8a", shape = 19) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_hline(yintercept = c(-1.96, 1.96)/sqrt(length(ts_data[[variable]])),
               linetype = "dashed", color = "red") +
    scale_x_continuous(breaks = seq(0, 36, by = 6)) +
    scale_y_continuous(limits = c(-1, 1)) +
    labs(title = paste("Autocorrelation Function (", title_prefix, ")"),
         x = "Lag", y = "Autocorrelation") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          panel.grid.minor = element_blank())

  # PACF Plot with identical styling
  pacf_data <- Pacf(ts_data[[variable]], lag.max = 36, plot = FALSE)
  pacf_df <- data.frame(lag = 1:length(pacf_data$acf),
                       PACF = as.numeric(pacf_data$acf))

  pacf_plot <- ggplot(pacf_df, aes(x = lag, y = PACF)) +
    geom_bar(stat = "identity", position = "identity",
             width = 0.2, fill = "#3f3f3f", color = NA) +
    geom_point(size = 3, color = "#8a8a8a", shape = 19) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_hline(yintercept = c(-1.96, 1.96)/sqrt(length(ts_data[[variable]])),
               linetype = "dashed", color = "#3639e6") +
    scale_x_continuous(breaks = seq(0, 36, by = 6)) +
    scale_y_continuous(limits = c(-0.8, 0.8)) +
    labs(title = paste("Partial Autocorrelation Function (", title_prefix, ")"),
         x = "Lag", y = "Partial Autocorrelation") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          panel.grid.minor = element_blank())

  # Print both plots (uncomment as needed)
  print(acf_plot)
  #print(pacf_plot)
}

# Generate ACF and PACF plots for all classifiers
A <- generate_acf_pacf("wa_cart_ha", "WA CART")
B <- generate_acf_pacf("wa_md_ha", "WA MD")
C <- generate_acf_pacf("wa_nb_ha", "WA NB")
D <- generate_acf_pacf("wa_rf_ha", "WA RF")

E <- generate_acf_pacf("precipitation_mm_mon", "Precipitation")
F <- generate_acf_pacf("Temperature", "Temperature")

combined_plot <- ((A | B) / (C | D) / (E | F)) + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(face = "bold"))
print(combined_plot)

ggsave("AF_combined_analysis.pdf", combined_plot,  width = 16, height = 12, dpi = 300)

```

## 4. Cross-Correlation with Precipitation and Temperature
Perform cross-correlation analysis between each classifier and precipitation/temperature.

```{r}
generate_ccf <- function(classifier, variable, title_suffix) {
  # Calculate CCF values
  ccf_data <- ccf(ts_data[[classifier]], ts_data[[variable]], lag.max = 12, plot = FALSE)
  ccf_df <- data.frame(lag = ccf_data$lag, CCF = ccf_data$acf)

  # Create plot with consistent styling
  ccf_plot <- ggplot(ccf_df, aes(x = lag, y = CCF)) +
    geom_bar(stat = "identity", position = "identity",
             width = 0.1, fill = "#3f3f3f", color = NA) +
    geom_point(size = 1.5, color = "#8a8a8a", shape = 19) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_hline(yintercept = c(-1.96, 1.96)/sqrt(ccf_data$n.used),
               linetype = "dashed", color = "#85002c") +
    scale_x_continuous(breaks = seq(-12, 12, by = 2)) +
    scale_y_continuous(limits = c(-1, 1)) +
    labs(title = paste(title_suffix),
         x = "Lag", y = "Cross-Correlation") +
    theme_bw(base_size = 10) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          panel.grid.minor = element_blank())

  print(ccf_plot)
}

# Generate CCF plots for all classifiers against precipitation and temperature
classifiers <- c("wa_cart_ha", "wa_md_ha", "wa_nb_ha", "wa_rf_ha")
variables <- c("precipitation_mm_mon", "Temperature")
titles <- c("Precipitation", "Temperature")
class_lab <- c("WA-CART", "WA-MD", "WA-NB", "WA-RF")

List <- list()
k = 1
J = 1
for (classifier in classifiers) {
  for (i in seq_along(variables)) {
    List[[k]] <- generate_ccf(classifier, variables[i], paste0(class_lab[J], " vs ",titles[i]))
    k = k + 1
  }
  J = J +1
}

combined_plot_ccf <- ((List[[1]] | List[[2]] | List[[3]] | List[[4]]) / (List[[5]] | List[[6]] | List[[7]] | List[[8]])) + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(face = "bold"))
print(combined_plot_ccf)

ggsave("Cross_Correlation.pdf", combined_plot_ccf,  width = 12, height = 6, dpi = 300)
```


## 5. Statistical Tests
Perform Augmented Dickey-Fuller (ADF) and Ljung-Box tests for stationarity and randomness.

```{r}
# Perform ADF and Ljung-Box tests for all classifiers
for (classifier in classifiers) {
  cat("\n--- Results for", classifier, "---\n")

  # ADF Test
  adf_test <- tseries::adf.test(ts_data[[classifier]])
  print(adf_test)

  # Ljung-Box Test
  ljung_box <- Box.test(ts_data[[classifier]], lag = 24, type = "Ljung-Box")
  print(ljung_box)
}
```

## 6. Seasonal Visualization
Visualize seasonal patterns for all classifiers.

```{r}
# Define a function for seasonal boxplots
generate_seasonal_plot <- function(classifier, title_suffix) {
  seasonal_plot <- ggplot(ts_data, aes(x = month(date), y = .data[[classifier]])) +
    geom_boxplot(aes(group = month(date))) +
    stat_smooth(method = "loess", se = FALSE) +
    theme_bw(base_size = 12) +
    labs(
      x = "Month",
      y = paste("Water Area (ha) - ", title_suffix, sep = ""),
      title = paste("Seasonal Pattern of Water Area (", title_suffix, ")", sep = "")
    )

  print(seasonal_plot)
}

List <- list()
titles <- c("WA-CART", "WA-MD", "WA-NB", "WA-RF")
k = 1
for (i in seq_along(classifiers)) {
    List[[k]] <- generate_seasonal_plot(classifier[i], titles[i])
    k = k + 1
}

combined_plot_ccf <- ((List[[1]] | List[[2]]) / (List[[3]] | List[[4]])) + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(face = "bold"))
print(combined_plot_ccf)

ggsave("Seasonal_Comb.pdf", combined_plot_ccf,  width = 6, height = 6, dpi = 300)

```

```{r}
library(ggplot2)
library(purrr)
library(patchwork)
library(lubridate)

# First calculate global y-axis limits across all classifiers
global_y_limits <- range(ts_data[, classifiers], na.rm = TRUE)

# Define a function for seasonal boxplots with fixed y-limits
generate_seasonal_plot <- function(classifier, title_suffix) {
  ggplot(ts_data, aes(x = factor(month(date)), y = .data[[classifier]])) +
    geom_boxplot(aes(group = month(date)), fill = "#e4e4e4", color = "#727272") +
    stat_smooth(aes(x = month(date)), method = "loess", se = FALSE, color = "#e64949", linewidth = 0.7) +
    theme_bw(base_size = 12) +
    labs(
      x = "Month",
      y = expression(paste("Water Area (ha)")),
      title = paste("Seasonal Pattern: ", title_suffix)
    ) +
    scale_x_discrete(labels = month.abb) +
    scale_y_continuous(limits = global_y_limits) +  # Apply global limits
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Classifiers and titles
classifiers <- c("wa_cart_ha", "wa_md_ha", "wa_nb_ha", "wa_rf_ha")
titles <- c("WA-CART", "WA-MD", "WA-NB", "WA-RF")

# Generate plots and store in list
plot_list <- map2(classifiers, titles, ~generate_seasonal_plot(.x, .y))

# Combine plots with patchwork
combined_plot <- (plot_list[[1]] | plot_list[[2]]) /
                 (plot_list[[3]] | plot_list[[4]]) +
                 plot_annotation(tag_levels = 'A') &
                 theme(plot.tag = element_text(face = "bold"))

# Save and display
print(combined_plot)
ggsave("Seasonal_Comb.pdf", combined_plot, width = 10, height = 8, dpi = 300)
```
